---
title: "Calibrating Interval Estimates using the Bootstrap"
format: r-wasm/live-revealjs
---

# Review

## Point and Interval Estimation

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}

```{webr}
#| label: ggplot-theme
#| include: false

lightgray = "#aaaaaa"
body.bg = rgb(231, 240, 236, maxColorValue=255)
gridcolor = rgb(0,0,0,.1, maxColorValue=1)

lab.theme = theme(
        plot.background = element_rect(fill = body.bg, colour = NA),
        panel.background = element_rect(fill = body.bg, colour = NA),
                    legend.background = element_rect(fill=body.bg, colour = NA),
                    legend.box.background = element_rect(fill=body.bg, colour = NA),
                    legend.key = element_rect(fill=body.bg, colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = lightgray),
        axis.text.y  = element_text(colour = lightgray),
        axis.title.x  = element_blank(),
        axis.title.y  = element_blank())

theme_set(lab.theme)
set.seed(1)
```


{{< include shared-code.qmd >}}
{{< include ga-turnout.qmd >}}

```{r}
#| fig-asp: .3 

set.seed(default.seed)
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(m, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
meanY.samples = rowMeans(Ys)
pmf.height = max(dbinom(0:n, n, mean(y))) 

scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025)) 
scale_y = scale_y_continuous(breaks=seq(0, pmf.height,by=.01))

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.8) + 
                      scale_x + scale_y
                      
bar.plot + with_annotations(meanY.samples, interval.width=width(meanY.samples), height=pmf.height) + 
           coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))
```

- In last week, we talked about estimating population frequencies.
- Our **point estimates** were frequencies in a random sample drawn from the population. 
- To characterize our uncertainty, we used **interval estimates**.
  - In particular, *calibrated* interval estimates. Confidence intervals.  
  - To get these intervals, we add 'arms' of equal length to our point estimate.
  - We choose the arm length so that, in 95% of surveys done like ours, the estimation target is within them.

## Calibration of Interval Estimates
```{r}
#| fig-asp: .3

bar.plot + with_annotations(meanY.samples, interval.width=width(meanY.samples), 
                                          shade.width=width(meanY.samples), height=pmf.height) +
           coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))
```

- We can do that using the **sampling distribution** of our point estimate.
- We'd draw arms out from the sampling distribution's center until they span 95% of point estimates.
- We can see that this is the right length by a shift of perspective.
  - A point estimate can touch the mean with its arms iff^[if and only if] the mean can touch it with equally long arms.
  - So 95% of intervals cover if and only if 95% of point estimates are within the mean's arms.
- In practice, we can't do exactly that. We don't know the sampling distribution.
- But we can use an estimate of the sampling distribution in its place. 
- If it's a good one, we'll get approximately 95% coverage.

## A Parametric Estimate of the Sampling Distribution 
```{r}
#| label: sampling-dist-comparison-binomial
#| fig-width: 7
#| fig-asp: .7
#| layout-ncol: 2 
#| fig-cap:
#| - Our point estimate and its sampling distribution
#| - Our estimate of its sampling distribution 

estimated.dist.samples = rbinom(10000, n, mean(Y)) / n 
bar.plot.estimated = ggplot() + 
  geom_bar(aes(x=estimated.dist.samples, y=after_stat(prop)), alpha=.3) +
  geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.7) + 
  scale_x + scale_y 

xx = seq(0,n)
p.parametric.truth = bar.plot + 
  with_annotations(meanY.samples, interval.width=width(meanY.samples), highlight.dots=1, height=pmf.height) +
  geom_line(aes(x=xx/n, y=dbinom(xx, n, mean(y))), color='black', alpha=.6) 
p.parametric.estimate = bar.plot.estimated  +
  with_annotations(c(mean(Y), estimated.dist.samples), highlight.dots=1, height=pmf.height,
                   shade.width=width(estimated.dist.samples), shade.color='red') +
  geom_line(aes(x=xx/n, y=dbinom(xx, n, mean(Y))), color='black') 

coord  = coord_flip(xlim=c(.6,.8), ylim=c(0, pmf.height))
p.parametric.truth + coord 
p.parametric.estimate + coord
```

 - Our estimate was based on knowledge of the **parametric form** of the sampling distribution.
 - When we sample with replacement, it's **binomial**.
    - It's the distribution of the frequency of heads in `r n` flips of a coin with probability $\theta$
    - where $\theta$ is the 'frequency of heads' in the population.
    - e.g. $\theta\approx `r approx(mean(y))`$ is the proportion of **registered voters** who will vote in our turnout example.

 $$
 P_\theta(\hat\theta = t) = \binom{n}{nt} \theta^{nt} (1-\theta)^{n(1-t)} \qqtext{ is the probability of the sample frequency being $t$}
 $$


- By *parametric form*, I mean a formula in terms of some parameters. 
  - That tells us what parameters we need to estimate to estimate the sampling distribution.
  - In this case, the only (unknown) parameter is the $\theta$, the frequency of heads in the population.
  - We plugged in the sample frequency, $\hat\theta$, to get our estimate of the sampling distribution.

$$
\hat P_\theta(\hat\theta = t) = \binom{n}{nt} \hat\theta^{nt} (1-\hat\theta)^{n(1-t)} \qqtext{ is our estimate of the same thing}
$$


## Sampling without Replacement

```{r}
#| label: sampling-dist-comparison-hypergeometric
#| fig-width: 7
#| fig-asp: .7
#| layout-ncol: 2 
#| fig-cap:
#| - Our point estimate and its sampling distribution
#| - Our estimate of its sampling distribution 

mh = 200
theta = 1/2
nh = 100
dist.samples = rhyper(10000, mh*theta, mh*(1-theta), nh)/ nh 
theta.hat = dist.samples[1]
estimated.dist.samples = rhyper(10000, round(mh*theta.hat), round(mh*(1-theta.hat)), nh) / nh

pmf.height.hyper = max(dhyper(0:n, mh*theta, mh*(1-theta), nh))
scale_x = scale_x_continuous(breaks=seq(.35,.65, by=.025))
scale_y = scale_y_continuous(breaks=seq(0, pmf.height.hyper,by=.01))
bar.plot.hyper = ggplot() + geom_bar(aes(x=dist.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=theta), color='green', size=1.5, alpha=.7) + 
                      scale_x + scale_y 

bar.plot.estimated = ggplot() + 
  geom_bar(aes(x=estimated.dist.samples, y=after_stat(prop)), alpha=.3) +
  geom_vline(aes(xintercept=theta), color='green', size=2, alpha=.7) + 
  scale_x + scale_y 

xx = seq(0,nh)
p.parametric.truth = bar.plot.hyper + 
  with_annotations(dist.samples, interval.width=width(dist.samples), highlight.dots=1, height=pmf.height.hyper) +
  geom_line(aes(x=xx/nh, y=dhyper(xx, mh*theta, mh*(1-theta), nh)), color='black', alpha=.6) 
p.parametric.estimate = bar.plot.estimated  +
  with_annotations(c(theta.hat, estimated.dist.samples), highlight.dots=1, height=pmf.height.hyper,
                   shade.width=width(estimated.dist.samples), shade.color='red') +
  geom_line(aes(x=xx/nh, y=dhyper(xx, mh*theta.hat, mh*(1-theta.hat), nh)), color='black') 

coord  = coord_flip(xlim=c(.35,.65), ylim=c(0, pmf.height.hyper))
p.parametric.truth + coord 
p.parametric.estimate + coord
```

- We can do the same thing when we sample *without replacement*.
- We just have to use a different parametric form: the **hypergeometric distribution**.

$$
 P_\theta(\hat\theta = t) =  \binom{n}{nt} \frac{\{m(1-\theta)\}!}{\{m(1-\theta)-n(1-t)\}!} \times \frac{(m\theta)!}{(m\theta-nt)!} \times \frac{(m-n)!}{m!} 
$$
 

- Again, the only unknown parameter is $\theta$. And we can plug in our point estimate to estimate this distribution.

$$
 \hat P_\theta(\hat\theta = t) =  \binom{n}{nt} \frac{\{m(1-\hat\theta)\}!}{\{m(1-\hat\theta)-n(1-t)\}!} \times \frac{(m\hat\theta)!}{(m\hat\theta-nt)!} \times \frac{(m-n)!}{m!} 
$$

# An Exercise

## A Survey
```{webr}
n=6
Y = c(0,0,0,1,1,1)

ggplot() + geom_point(aes(x=1:6, y=Y))
```

::: {.panel-tabset}
### The Idea
- We're going to run a survey on Candy Preferences.
  - I'll draw a sample of size $n=6$, with replacement, from the people in the room.
  - They'll pick candy and write their selection into a sample table on the board.
  - In particular, we'll write out whether they chose pickle ($Y=1$) or watermelon ($Y=0$) sour straws.
- Then we're going to estimate the sampling distribution in a new way.
  - It'll be a lot like calculating the actual sampling distribution.
  - We'll draw a sample of size $n$ with replacement, use it to calculate our estimator, and repeat.
- What's different is the population we're drawing our sample from.
- We don't have responses from the whole population, so we'll draw it from the closest thing we've got: the sample.
  - We call this *bootstrapping*. 
  - The estimate of the sampling distribution we get is called the *bootstrap sampling distribution*.

### Instructions

- I've drawn the sample we got from our survey on the board.
- Here's how you draw a sample from the bootstrap sampling distribution of our point estimator.
  1. Roll your die 6 times to draw a sample of size $n=6$ *from the sample*.
  2. Calculate the sample frequency. That's one draw from the bootstrap sampling distribution. Write it down.
- If we all do this 5 times, we'll have a bunch of draws.
- We'll tally them up on the board and visualize the result as a histogram.
- A histogram of draws from the *bootstrap sampling distribution*.

:::

## Simulating a Few More Draws

```{webr}
draws = 1:10000 |> map_vec(function(.) { 
  Jstar = sample(1:n, n, replace=TRUE)
  Ystar = Y[Jstar]
  mean(Ystar)
})
computer.tally = as.data.frame(table(draws)) |>
  rename(theta.hat=draws, count=Freq)
```

## Our Tally

```{webr}
hand.tally = data.frame(theta.hat = c(0,1,2,3,4,5,6),
                        count     = c(0,0,0,1,0,0,0))

ggplot() +
  geom_col(aes(x=(0:n)/n, y=hand.tally$count/sum(hand.tally$count)),
           fill='red', color='red', linewidth=.1, alpha=.2)
```

- Type the room's tally into `count` above and run it.
- Each entry is the number of our 10 bootstrap draws that came out to that $\hat\theta$.

## Compared to the Binomial Estimate

```{webr}
ggplot() + geom_col(aes(x=(0:n)/n, y=dbinom(0:n, n, mean(Y))), color='blue', fill='blue', linewidth=.1, alpha=.2) +
           geom_col(aes(x=(0:n)/n, y=hand.tally$count/sum(hand.tally$count)), fill='red', color='red', linewidth=.1, alpha=.2)
```

::: {.panel-tabset}
### Success!
- If all has gone to plan, [our histogram]{.fg style="color:red"} looks a lot like our [binomial estimate]{.fg style="color:blue"} of the sampling distribution.
  - If we substitute in a computer tally of 10,000 draws from the bootstrap sampling distribution, we nail it.
  - It looks like the bootstrap sampling distribution *is* the binomial estimate. 
- **Q**. It is. How do you know?

### A Big Hint.
  - We know the distribution of the frequency of 1s in a sample of size $n$ drawn with replacement ...
    - ... from a population $y_1\ldots y_m$ of binary responses with frequency $\theta$.
  - It's $\text{Binomial}(n, \theta)$. To estimate it, we plug in $\hat\theta$, the frequency of 1s in our sample. 
  - So our Binomial estimate is the distribution of the frequency of 1s in a sample of size $n$ drawn with replacement ...
    - ... from 'a population' $Y_1 \ldots Y_n$ with frequency $\hat\theta$.
  - What's a draw from the bootstrap sampling distribution?

### Answer.

- Each bootstrap sample is the frequency of 1s in a sample of size $n$ drawn with replacement ...
  - ... from  'a population' $Y_1 \ldots Y_n$ in which the frequency of 1s is $\hat\theta$,  the frequency of 1s in our sample. 
  - Because that 'population' is our sample.

:::

# The Bootstrap

## The Bootstrap Interpretation in Our Turnout Poll
```{r}
#| fig-asp: .8 
#| fig-width: 4.5
#| layout-ncol: 3

boot.ind = sample(1:n, n, replace=TRUE)

displaypop = pop[1:50000,]
pop$sam.replicated   = as.numeric(table(c(1:m, Js[1,])) - 1)
pop$boot.replicated  = as.numeric(table(c(1:m, Js[1, boot.ind])) - 1)

pop.map  = state.plot + geom_point(aes(x = long, y = lat), 
                                   alpha=.05, data=displaypop, size=.025, color=polla) + 
                        scale_size_area() + guides(size='none')
sam.map  = state.plot + geom_point(aes(x = long, y = lat, size=sam.replicated),   
                                   alpha=.2, data=pop[pop$sam.replicated > 0, ], color=polla) + 
                        scale_size_area(max_size=3) + guides(size='none')
boot.map = state.plot + geom_point(aes(x = long, y = lat, size=boot.replicated),   
                                   alpha=.2, data=pop[pop$boot.replicated > 0, ], color=polla)   +
                        scale_size_area(max_size=6) + guides(size='none')

pop.map
sam.map
boot.map
```


:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"}

- The sampling distribution estimate we've used was $\text{Binomial}(n,\hat\theta)$ for $n=`r n`$ and $\hat\theta \approx `r approx(mean(Y))`$.
  - It's the distribution of the proportion heads in `r n` flips of a coin with probability $\hat\theta \approx `r approx(mean(Y))`$ of heads.
  - where $\hat\theta \approx `r approx(mean(Y))`$ is the proportion of **voters we've polled** who will vote. 
- That is, it's the sampling distribution of a 'poll' of the people in our sample, i.e.
    - roll a 625-sided die 625 times
    - call up the corresponding person in our sample
    - and counting up the yeses we hear
- Note that this is random because we're drawing with replacement.
  - Each time we run this poll, we call each person in our sample 0,1,2,... times
  - And the number of times we call them is random.

:::
::: {.fragment .current-visible data-fragment-index="0"}

- If we plot our voters on a map, you can see the idea in visual terms.
    - On the left, we have the population. 
    - In the middle, we have our sample. It's drawn, with replacement, from the population.
    - On the right, we have something else. A new sample drawn, with replacement, *from the sample*.
- Each 'call' that a person receives increases the size of their dot: $\text{circle area} \propto \text{number of calls}$.
- In the sample, even though it's drawn with replacement, all dots are the same size.
  - Because we draw from such a large population, nobody gets called twice.
- In the bootstrap sample, dots vary in size.
  - Because we draw $n$ people from a sample of size $n$, it's almost impossible not to call somebody twice. 

:::
::::

## Bootstrapping

- Before the election, we don't observe the population. But we do observe the sample. 
  - And we can *sample from the sample* and act as if it were the population.
  - We'll take repeated random samples of size `r n`, with replacement, from our sample of size `r n`.
- We call these *bootstrap samples* and estimates based on them *bootstrap estimates*.
  - The distribution of these estimates is called the *bootstrap sampling distribution*.
  - If the sample is like the population, this should be like our estimator's actual sampling distribution.

```{r bootstrap.sample}
J.star = sample(n, size=n, replace=TRUE)
Y.star = Y[J.star]
```

:::: columns
::: column
The Sample
$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

The Bootstrap Sample

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}^*  \\
Y_i^*  &  `r Y.star[1]` & `r Y.star[2]` & \dots & `r Y.star[n]` & `r approx(mean(Y.star))` \\
\end{array}
$$
:::

::: column
The Population

$$
\begin{array}{r|rrrr|r}
j & 1 & 2 & \dots &   `r approx(m)` & \bar{y}_{`r approx(m)`}  \\
y_{j} & `r y[1]` & `r y[2]` & \dots & `r y[m]`   & `r approx(mean(y))`  \\
\end{array}
$$

The 'Bootstrap Population' --- The Sample
$$
\begin{array}{r|rrrr|r}
j        &   1 & 2 & \dots & `r n` & \bar{y}^*_{`r n`}  \\
y_j^*  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::::

::: aside
Notation. We use stars to distinguish the bootstrap sample from our original sample---we write $Y_i^*$ and $\bar Y^*$.
:::

## The Bootstrap is *Nonparametric* 

```{r}
#| echo: true

bootstrap.samples = array(dim=10000)
for(rr in 1:10000) {
    Y.star = Y[sample(1:n, n, replace=TRUE)]
    bootstrap.samples[rr] = mean(Y.star) 
}
```

```{r}
#| out-height: 400px 
#| fig-asp: .3 
ggplot() + geom_bar(aes(x=bootstrap.samples, y=after_stat(prop)), alpha=.3, position='identity') + 
           geom_line(aes(x=(0:n)/n, y=dbinom(0:n, n, mean(Y))), color='black') + 
           xlim(.6,.8)
```

- We do not need to know the parametric form of our sampling distribution to use it.
- All we do is re-run our poll *acting as if* our sample were the population.
  - We can do this no matter what we're estimating. 
  - So let's try it out on some stuff where we don't know the parametric form.
  
  
# Beyond Binary Responses

## Vote History

```{r}
#| label: vote-history-setup

# Generate vote history: number of elections voted in out of the last 5
# Correlated with current turnout intention
set.seed(42)
# For the population: people who will vote (y=1) tend to have higher vote history
h = rep(NA, m)
h[y == 1] = rbinom(sum(y == 1), size = 5, prob = 0.7)  # voters average ~3.5
h[y == 0] = rbinom(sum(y == 0), size = 5, prob = 0.3)  # non-voters average ~1.5
pop$h = h

# Sample vote history
H = h[J1]
sam$h = H
```

- So far we've asked a binary question: will you vote?
- But voter files contain richer information.
    - For each registered voter, we can see how many of the last 5 elections they voted in.
    - Let's call this *vote history*, and write it $H_i$ in our sample and $h_j$ in the population.

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{H}_{`r n`}  \\
H_i  &  `r H[1]` & `r H[2]` & \dots & `r H[n]` & `r round(mean(H), 2)` \\
\end{array}
$$

- We might want to estimate the average vote history in the population.
    - Our point estimate would be the sample mean, $\bar H_n = `r round(mean(H), 2)`$.
    - But what about the sampling distribution?

## No Parametric Form This Time

- When the response was binary, we knew the parametric form: Binomial.
    - The mean of $n$ binary responses, each $1$ with probability $\theta$, has a $\text{Binomial}(n, \theta)/n$ distribution.
- But now our responses take values in $\{0, 1, 2, 3, 4, 5\}$.
    - There's no simple parametric form for the distribution of their mean.
- We could work one out.
    - A response $H_i$ has some probability $p_0$ of being 0, $p_1$ of being 1, and so on.
    - The sum $H_1 + \ldots + H_n$ is a sum of $n$ such random variables, and its distribution could be computed by convolution.
    - But that's a pain. And it'd be an even bigger pain if our responses were continuous.
- The bootstrap gives us a way out. We don't need to know the parametric form.
    - We just resample from the sample.

## The Sampling Distribution of Average Vote History

```{r}
#| echo: true
#| eval: false
mean.history.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:m, n, replace=TRUE)
    Hsam = h[I]
    mean.history.samples[rr] = mean(Hsam)
}
```

```{r}
set.seed(default.seed)
mean.history.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:m, n, replace=TRUE)
    Hsam = h[I]
    mean.history.samples[rr] = mean(Hsam)
}
```

```{r}
#| fig-asp: .4
history.histogram = ggplot() +
  geom_histogram(aes(x=mean.history.samples, y=after_stat(density)), bins=50, alpha=.3)
history.histogram +
  with_annotations(mean.history.samples,
                   shade.width=width(mean.history.samples),
                   interval.width=width(mean.history.samples),
                   estimand=mean(h)) +
  scale_x_continuous(limits=c(1.8, 2.8))
```

- That's what the *sampling distribution* of average vote history looks like.
- As before, if we can estimate it we can use that estimate to get a 95% confidence interval.
- But, unlike before, we don't have a nice parametric form like the Binomial.
    - So we'll use the bootstrap to estimate it.

## The Bootstrap Sampling Distribution of Average Vote History

```{r}
#| echo: true

history.bootstrap.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    Hstar = H[I]
    history.bootstrap.samples[rr] = mean(Hstar)
}
```

```{r}
#| fig-width: 7
#| fig-asp: .5
#| layout-ncol: 2
#| fig-cap:
#| - The actual sampling distribution
#| - The bootstrap sampling distribution

scale_x_hist = scale_x_continuous(limits=c(1.8, 2.8))
w.hist = width(mean.history.samples, alpha=.05)
w.hist.boot = width(history.bootstrap.samples, alpha=.05)

bootstrap.history.histogram = ggplot() +
  geom_histogram(aes(x=history.bootstrap.samples, y=after_stat(density)), alpha=.3, bins=50)

history.histogram +
  with_annotations(mean.history.samples, interval.width=w.hist.boot, highlight.dots=1,
                   estimand=mean(h)) +
  coord_flip() + scale_x_hist

bootstrap.history.histogram +
  with_annotations(c(mean(H), history.bootstrap.samples),
                   highlight.dots=1, shade.width=w.hist.boot, shade.color='red') +
  coord_flip() + scale_x_hist
```

- It looks like it works in this case.
    - The bootstrap sampling distribution is a good approximation to the actual sampling distribution.
    - And the interval widths are similar.

## But We Can't Argue That It Should

- We're no longer able to argue that it *should* work the way we did before.
    - For the binary case, we took advantage of our knowledge of the parametric form.
    - We knew the bootstrap sampling distribution *was* the plug-in binomial estimate.
    - And we don't have that now.
- We'll get there. But we'll need a few new tools we'll develop in the coming weeks.
    - **Normal approximation**, which gives us a parametric form for an *approximation* to our estimator's sampling distribution.
    - **Techniques for variance calculation**, which help us understand the parameters that go into it.
